---
title: "DS 6372 EDA"
author: "Oluwadamilola Owolabi"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---
title: "Budwiser EDA"
authors: "Anthony and Damilola"
date: "02/27/2024"
output: html_document
editor_options: 
  chunk_output_type: console
—


Introduction: 
Good afternoon Mr. Whitworth and Mr. Tennenbaum. We are here today to present the findings from the questions presented to our team of Data Scientist. We will answer the nine questions provided ahead of time with our analysis and visualuations, as well as highlighting any key observations that came up during our review. With that, let us get started as we’re excited to share with you the breakdown of our findings. 


Installing the packages

```{r}
#install.packages("tidyverse")
#install.packages("ggplot2")
#install.packages("GGally")
#install.packages("dplyr")
#install.packages("maps")
#install.packages("mapsproj")
library(ggplot2)
library(tidyverse)
library(GGally)
library(dplyr)
library(class)
library(caret)
library(e1071)
library(maps)
library(mapproj)

```

#### Agenda 1a: Distribution by State - Load files and confirm, how many breweries are located in each state?

```{r}
#Load file #1 Beers.csv
Beers = read.csv(file.choose(),header = TRUE) #load the first file
head(Beers) #confirm the file is correct, review 

#Load file #2 Breweries.csv
Breweries = read.csv(file.choose(),header = TRUE) #load the second file
head(Breweries) #confirm the file is correct, review

summary(Breweries) #check the type

mean_count <- mean(Breweries$n) #create a column mean_count for the reference line
    
Breweries %>% group_by(State) %>% summarize(n = n()) %>% ggplot(aes(x = State, y = n, fill = State)) +
  geom_bar(stat = "identity") + theme(legend.position = "none") + geom_text(aes(label = n, y = n +1.5)) +
  ggtitle("Count of Breweries by State") + ylab("Count of Breweries") + xlab("State") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  geom_hline(yintercept = mean_count, linetype = "dotted", color = "red") # to show the mean.

```
#### Answer 1a - We provided a bar chart showing the count of breweries by state. The state with the highest count of breweries is Colorado with 47 breweries. There is a four way tie between DC, North Dakota, South Dakota, and West Virginia for the lowest with one brewery each. The mean count of breweries per state is eleven. 

#### Agenda 1b: Distribution by State - Load files and confirm, how many breweries are located in each state?

```{r}
lookup = data.frame(State = state.abb, State_Name = state.name) #makes a dataframe with State name and state abb. 
head(lookup)

Breweries2 = inner_join(Breweries,lookup, by = "State")
head(Breweries2)
BreweriesMapData = Breweries2 %>% group_by(State) %>% count(State_Name)
head(BreweriesMapData)
colnames(BreweriesMapData)[3] = "Breweries" #change 'n' to 'Breweries'
BreweriesMapData$region <- tolower(BreweriesMapData$State_Name)
head(BreweriesMapData)
BreweriesMapData2 = BreweriesMapData[-1] #drop the first column
states <- map_data("state")
head(states)
map.df <- merge(states, BreweriesMapData2, by = "region", all.x = TRUE)
head(map.df)
map.df <-map.df[order(map.df$order),]
ggplot(map.df,aes(x = long, y = lat, group = group))+
         geom_polygon(aes(fill = Breweries)) + 
         geom_path() +
         scale_fill_gradientn(colours=rev(heat.colors(10)), na.value = "grey90") + ggtitle("Breweries by State") + coord_map() 


```
#### Answer 1b/9 - In addition to the bar chart, a heat map showing the distribution of breweries was included. This was added as a “knock their socks off” addition as it enhanced the answer to their question visually. 

#### Question 2 - Merge beer data with the breweries data. Print the first 6 observations and the last six observations to check the merged file.  

```{r}
merged_df = merge(Beers,Breweries, by.x = "Brewery_id", by.y = "Brew_ID") #joining when keys do not match
merged_df 
head(merged_df,6) #to show the first 6 observations 
tail(merged_df,6) #to show the last 6 observations
#suggestion for enhancement - rename name.x to beer_name, rename name.y to brewery_name
```
#### Answer 2 - Using merge() we brought in the Breweries csv file against the Beers csv file. The method used was when the two keys were not identical.

#### Agenda 2: Missing Values - Address the missing values in each column

```{r}
merged_df %>% is.na() #check for missing values , ABV and IBU

#Cleaning the data
merged_df <- merged_df %>% na.omit() 

merged_df %>% is.na() #check for missing values , ABV and IBU
merged_dfna = sapply(merged_df, function(x) sum(is.na(x)))
merged_dfna #confirmed across all columns, that only ABV an IBU have n/a values
summary(merged_df) #verify type

#confirming the presence os NAs
#check_for_na <- function(data) {
 # if (any(is.na(data))) {
#    return("Data contains missing values (NAs).")
#  } else {
#    return("Data does not contain missing values (NAs).")
#  }
#}

#check_for_na(merged_df)

```
#### Answer 3 - There were 62 missing values from column ABV, approximately three percent of the total. There were 1,005 missing values from column IBU. The data was review and we could not locate a relationship amongst the missing data and the qualifiers. Example, there did not appear to be an observable difference between data missing across the State data, or Brewery data. Overall, it appeared that the data was missing completely at random with no way of back tracking into an answer for the missing values. 

#### Agenda 3: Mean ABV %  - mean of ABV by State and bar chart

```{r}
#create a new df to add Mean ABV % 
merged_df1 <- merged_df %>% group_by(State) %>% summarize(meanABV = mean(ABV, na.rm = TRUE)) #add column meanABV with the mean of ABV. 
#confirm meanABV has been added 
merged_df1
#plot to show the meanABV % by state, update labels, remove legend. 
ggplot(merged_df1, aes(x = State, y = meanABV, fill = State)) + 
  theme(legend.position = "none") + #removed the legend
  geom_bar(stat = "identity") + geom_text(aes(label = round(meanABV, 2), y = meanABV + .01), size = 2) + #using round for cleaner looking labels
  ggtitle("ABV % mean by State") + xlab("State") + ylab("Mean ABV %") + 
  ylim(0.0,0.1) + #zoom in on the y-axis for small %s
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```
#### Answer 4a - New Mexico has the highest mean Alcohol by Volumne % (ABV %) and Utah has the lowest ABV %. Each state was calculated and a label was added to show each. To prevent overlapping, the size of the mean labels were reduced.  

#### Agenda 4: Mean IBU  - mean of IBU by State and bar chart

```{r}
#repeat with IBU
#create a new dataframe with 
merged_df2 <- merged_df %>% group_by(State) %>% summarize(meanIBU = mean(IBU, na.rm = TRUE)) #add column meanABV with the mean of IBU. 
#confirm meanIBU has been added 
merged_df2
#plot to show the meanIBU by state, update labels, remove legend.
ggplot(merged_df2, aes(x = State, y = meanIBU, fill = State)) + 
  theme(legend.position = "none") + #removed the legend
  geom_bar(stat = "identity") + geom_text(aes(label = round(meanIBU, 1), y = meanIBU + 2), size = 2) + #using round for cleaner looking labels
  ggtitle("IBU mean by State") + xlab("State") + ylab("Mean IBU") + 
  ylim(0,60) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

#### Answer 4b - West Virginia has the highest mean International Bitterness Units (IBU). Due to missing values, South Dakota had no data represented. The remaining states with data were calculated and a label was added to show each. To prevent overlapping, the size of the mean labels were reduced.  


#### Agenda 5 - Finding the states with the most alcoholic beer and the most bitter beer.

```{r}

#Which state has the maximum alcoholic (ABV) beer 
merged_df %>%  filter(!is.na(ABV)) 
maxABV <- max(merged_df[["ABV"]]) #getting the max ABV value
merged_df %>% group_by(State) %>% ggplot(aes(x = State, y = ABV, fill = State)) + theme(legend.position = "none") + geom_boxplot()  + ggtitle("Stem and Box plot of the ABV Values for each state")

#Which state has the most bitter (IBU) beer?
MostBitterIBU <- max(merged_df[["IBU"]]) #getting the max IBU value, the higher the IBU, the more bitter the beer
MostBitterIBU
MostBitterIBU_state <- merged_df[grep(MostBitterIBU, merged_df$IBU), "State"] #finding the state for the max IBU value
MostBitterIBU_state

merged_df %>% group_by(State) %>% ggplot(aes(x = State, y = IBU, fill = State)) + theme(legend.position = "none") + geom_boxplot()  + ggtitle("Stem and Box plot of the IBU Values for each state")
```

#### Results: The state of Kentucky possesses the most alcoholic beverage with an ABV of 12.5%, while the state of Oregon, possesses the most biter beer with an IBU of 138 units

#### Agenda 6: Investigating the summary statistics of the Alcohol By Volume Variable.

```{r}

#Getting the summary
summary(merged_df$ABV)

MaxABV <- max(merged_df[["ABV"]])
MinABV <- min(merged_df[["ABV"]])

#Getting a  box plot for the summary
merged_df %>% group_by(ABV) %>% ggplot(aes(y = ABV)) + geom_boxplot(color = 'orange') + ggtitle("The summary statistics of the ABV variable") + theme(legend.position = "none")

#Visualizing the distribution based on the states using a box plot

#Getting the boxplot for the state with the highest ABV (Kentucky)
merged_df[grep("KY", merged_df$State), ] %>% group_by(State) %>% ggplot(aes(y = ABV, x = State, fill = State)) + geom_boxplot() + ggtitle("The distribution of the ABV based on the state") + theme(legend.position = "none")

#Getting the boxplot for all states
merged_df %>% group_by(State) %>% ggplot(aes(y = ABV, x = State, fill = State)) + geom_boxplot() + ggtitle("The distribution of the ABV based on the state") + theme(legend.position = "none")

MaxABV; merged_df[grep(MaxABV, merged_df$ABV), "State"] #Getting the state of MaxABV
MinABV; merged_df[grep(MinABV, merged_df$ABV), "State"] #Getting the state of MinABV

#Visualizing the distribution based on the city using a barchart
merged_df %>% group_by(City) %>% ggplot(aes(y = merged_df$ABV, x = City, fill = State)) + geom_boxplot() + ggtitle("The distribution of the ABV based on the City")

#Visualizing the distribution based on the cities in Kentucky using a barchart
merged_df[grep("KY", merged_df$State), ] %>% ggplot(aes(y = ABV, x = City, fill = City)) + geom_boxplot() + ggtitle("The distribution of the ABV based on cities in Kentucky") + theme(legend.position = "none")

#Visualizing the distribution based on the Brewery ID using a scatterplot
merged_df %>% group_by(Brewery_id) %>% ggplot(aes(y = ABV, x = Brewery_id, fill = Brewery_id)) + geom_point() + ggtitle("The distribution of the ABV based on each Brewery ID")

MaxABV; merged_df[grep(MaxABV, merged_df$ABV), "Name.x"] #Getting the Brewery ID of MaxABV
MinABV; merged_df[grep(MinABV, merged_df$ABV), "Name.x"] #Getting the Brewery ID of MinABV

#Visualizing the distribution based on the weight in ounces using a barplot
merged_df %>% group_by(Ounces) %>% ggplot(aes(y = ABV, x = Ounces, fill = Ounces)) + geom_bar(stat = "identity") + ggtitle("The distribution of the ABV based on the weight of the beer drink (in ounces")
# The bar just shows the frequency distribution of the ABV for each weight 

#Visualizing the distribution based on the IBU using a scatterplot
merged_df %>% group_by(IBU) %>% ggplot(aes(y = ABV, x = IBU, color = State)) + geom_point() + ggtitle("The distribution of the ABV based on each IBU") + theme(legend.position = "none")
                                                      
```
#### Results: Looking at the summary statistics, the minimum value is 0.027, the 1st quartile value is 0.05, the Median value is 0.057, the 3rd Quarter is 0.068, the maximum value is 0.125. Valuable insight were also gotten from different categorical variables related to the ABV variable. Looking at the distribution based on the states, the state of Kentucky has the highest ABV value, the state of Oregon has the lowest ABV values, and there were 2 outliers in the plot, belonging the the states of Kentucky and Oregon. Looking at the distribution based on the brewery ID, the Beer "London Balling" with brewery ID #2 has the highest alcoholic content, while the beer "Totally Randler" with brewery ID# 81 has the lowest alcoholic content.


#### Agenda 6.5 : Investigating the summary statistics of the International Business Unit Variable.

```{r}

#Getting the summary
summary(merged_df$IBU)

MaxIBU <- max(merged_df[["IBU"]])
MinIBU <- min(merged_df[["IBU"]])

#Getting a  box plot for the summary
merged_df %>% group_by(IBU) %>% ggplot(aes(y = IBU)) + geom_boxplot(color = 'brown') + ggtitle("The summary statistics of the IBU variable") + theme(legend.position = "none")


#Visualizing the distribution based on the states using a box plot
merged_df %>% group_by(State) %>% ggplot(aes(y = IBU, x = State, fill = State)) + geom_boxplot() + ggtitle("The distribution of the IBU based on the state") + theme(legend.position = "none")

MaxIBU; merged_df[grep(MaxIBU, merged_df$IBU), "State"] #Getting the state of MaxIBU
MinIBU; merged_df[grep(MinIBU, merged_df$IBU), "State"] #Getting the state of MinIBU

#Visualizing the distribution based on the city using a barchart
merged_df %>% group_by(City) %>% ggplot(aes(y = merged_df$IBU, x = City, fill = State)) + geom_boxplot() + ggtitle("The distribution of the IBU based on the City")  + theme(legend.position = "none")

#Visualizing the distribution based on the cities in Kentucky using a barchart
merged_df[grep("KY", merged_df$State), ] %>% ggplot(aes(y = IBU, x = City, fill = City)) + geom_boxplot() + ggtitle("The distribution of the IBU based on cities in Kentucky")

#Visualizing the distribution based on the Brewery ID using a scatterplot
merged_df %>% group_by(Brewery_id) %>% ggplot(aes(y = IBU, x = Brewery_id, fill = Brewery_id)) + geom_point() + ggtitle("The distribution of the IBU based on each Brewery ID")

MaxABV; merged_df[grep(MaxIBU, merged_df$IBU), "Name.x"] #Getting the Brewery ID of MaxIBU
MinABV; merged_df[grep(MinIBU, merged_df$IBU), "Name.x"] #Getting the Brewery ID of MinIBU

#Visualizing the distribution based on the weight in ounces using a barplot
merged_df %>% group_by(Ounces) %>% ggplot(aes(y = IBU, x = Ounces, fill = Ounces)) + geom_bar(stat = "identity") + ggtitle("The distribution of the IBU based on the weight of the beer drink (in ounces")
# The bar just shows the frequency distribution of the IBU for each weight 

#Visualizing the distribution based on the IBU using a scatterplot
merged_df %>% group_by(IBU) %>% ggplot(aes(y = ABV, x = IBU, color = State)) + geom_point() + ggtitle("The distribution of the ABV based on each IBU")  + theme(legend.position = "none")
                                                      
```
#### Results: Looking at the summary statistics, the minimum value is 4, the 1st quartile value is 21, the Median value is 35, the 3rd Quarter is 64, the maximum value is 138. Valuable insight were also gotten from different categorical variables related to the ABV variable. Looking at the distribution based on the states, the state of Oregon has the highest ABV value, and the state of Carlifornia has the lowest ABV values. Looking at the distribution based on the brewery ID, the Beer "Bitter B***th Imperial IPA" with brewery ID #2 has the highest alcoholic content, while the beer "Summer Solstice" with brewery ID# 81 has the lowest alcoholic content.



#### Agenda 7: Investigating the relationship between the ABV variable and IBU variable

```{r}

#Visualizing the distribution based on the ABV vs. IBU using a scatterplot
merged_df %>% group_by(IBU) %>% ggplot(aes(y = ABV, x = IBU, color = State)) + geom_point() + ggtitle("The distribution of the ABV based on each IBU") + theme(legend.position = "none")

```
#### Results: There is a positiver linear relationship between the bitterness of the beer and its alcoholic content. We can see  from  the plot above that for every high value in IBU there is a corresponding high value in ABV


#### Agenda 8: Investigating the difference between the IPAs (India Pale Ales) and other types of ales in terms of their IPA and IBU values using KNN classification. We shall provide statistical evidence using a 2 sample t-test for such investigation

```{r}

# Using KNNs
set.seed(123)

# Step 1: Data Preprocessing

#Extracting relevant data for the knn
Budweiser_df = merged_df %>% filter(str_detect(Style, "IPA") | (str_detect(Style, "Ale"))) #filtering the IPA and different ale beers using str_filter() & str_detect

#grouping all IPA and Ale variables to groupnames
Budweiser_df2 <- Budweiser_df %>%
  mutate(Style = case_when(
    grepl("IPA", Style, ignore.case = TRUE) ~ "IPA",
    grepl("Ale", Style, ignore.case = TRUE) ~ "Ale",
    TRUE ~ Style  # Keep the original name if no match
  ))

# Step 2: Data Exploration

#Visualizing the distribution of IBU and ABV values for the IPA and beers using a box plot
Budweiser_df2 %>% group_by(IBU) %>% ggplot(aes(y = IBU)) + geom_boxplot(color = 'violet') + ggtitle("The summary statistics of the IBU variable") + theme(legend.position = "none")

Budweiser_df2 %>% group_by(ABV) %>% ggplot(aes(y = ABV)) + geom_boxplot(color = 'purple') + ggtitle("The summary statistics of the ABV variable") + theme(legend.position = "none")

Budweiser_df2 %>% ggplot(aes(x = IBU, y = ABV,color = Style)) + geom_point() +  ggtitle("Distribution of ABV vs. IBU") 

# Step 3: Data Splitting
#sorting the datasets into training and test data, Using an 80-20 split
split_perc <- 0.80
trainPosition = sample(1:nrow(Budweiser_df2)[1],round(split_perc * nrow(Budweiser_df2)[1])) #getting the index for the training dataset
trainData <- Budweiser_df2[trainPosition, ]
testData <- Budweiser_df2[-trainPosition, ]
trainData; testData

#testing the dimensions of the train and test data

# Check for NA values
print(colSums(is.na(trainData)))
print(colSums(is.na(testData)))

# Check data types
print(sapply(trainData[, c("IBU", "ABV")], class))
print(sapply(testData[, c("IBU", "ABV")], class))

# Step 4: Building the KNN model
classifications <- knn(trainData[, c("IBU", "ABV")], testData[, c("IBU", "ABV")], trainData$Style, prob = TRUE, k = 5)
#table(classifications, testData$Style)

# Step 5: Model Evaluation
CM = confusionMatrix(table(classifications,testData$Style))
CM

#Statistical Evidence

#Using ttest to compare the mean of IBU and ABV variables
Budweiser_IPA = Budweiser_df2 %>% filter(str_detect(Style, "IPA"))
Budweiser_Ale = Budweiser_df2 %>% filter(str_detect(Style, "Ale"))

t.test(Budweiser_IPA$IBU, Budweiser_Ale$IBU)
t.test(Budweiser_IPA$ABV, Budweiser_Ale$ABV)
                                          
```
#### Results: Using the KNN classificayion, the accuracy was found to be 86%, the Missclassification rate 14%, sensitivity, 90.4%, and the specitivity, 79.8%. Based on the welch's 2 sample t-test, there is overwhelming evidence to suggest that there is a significant difference between the Ale and IPA drinks in term of their IBU (p-value < 2.2e-16) and the ABV (P-value < 2.2e-16).


Conclusion

Utilizing various file manipulation, string manipulation and data visualization techniques, we were able to find valuable insight into the data of US craft beers and US breweries, over the past years.

We were able to discover the distribution of breweries and beers, across different US states. We were also able to address missing values within our generated data, and investigated the effectiveness of the missing values within the dataset. From our investigation, we proved that Removing the missing values didnt affect our data. Statistical analysis was done to find and visualize valuable statistical data like states that produced drinks with the highest average values of the Alcohol by Volume variable and the International Bitterness Unit Variable, the stateswith most alcoholic drink and the most bitter drink. We were able to provide more statistical insight into the Alcohol by beverage data, by looking into their mean, maximum and minimum values, and looking into how they affect different variables like the IBU, the brewery ID, and the states.

Lastly we were able to use  kNN, and a ttest to prove that there is a significant relationship between the IPA drinks and the Ale drinks.



